Relational programming in Smalltalk

### 2019-03-19 booklet/article.tex
Update and apply suggestions by Stef Ducasse.

\sd{Why don't you start with Cons. Then once people understand Cons and infinite stream
you can introduce Goal and state and always example first.}

@@

\subsection{\ct{Goal}s and \ct{State}s}

A \textit{goal} is any object that responds to the message \ct{\#onState:} so
that it receives a substitution and produces an ordered sequence of zero or
more substitutions.

\sd{Show a test. I do not get what is a state? A variable? Why this is call onState:}
\sd{Why this is not \ct{goal withSubstitution: aSubstitution}}

 In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\ct{Dictionary} object fits perfectly \sd{show an example now}; we wrap it \sd{who?} in a \ct{State} object
that also keeps record of an incremental numbering  \ct{birthdate} whose
meaning will be clarified later when we'll describe the  \ct{Fresh}
goal\footnote{put here the reference to its subsection}. An empty substitution
is created by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary new asState
\end{minted}
where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary >> asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> at: aVar put: aValue
    | s |
    s := substitution copy.
    "check that if a key already exists at aVar, then the values must be the same"
    s 
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
maintaining a coherent protocol with respect to the one of \Verb|Dictionary|
objects for extension; moreover, \ct{UnificationError} objects will be
explained in a later section when will talk about the unification process.

\subsection{\ct{Chain}s and \ct{Cons}es}
\sd{As a general comments. You should say to the reader. We will do that. And do piece by piece
Now I discover things just stumbling on them.}

\sd{Example: In this section we will build the infrastructure to represent potentially infinite lists. 
For this will start building a chain. A chain is composed of consed: objects following Lisp tradition. 
We will define first the class Chain as an abstract class with two subclass \ct{Knot} and \ct{Bottom}.
\ct{Bottom} will represent the empty cell of lisp on top of which any list is build. \ct{Knot} represents 
a ...
}

Because the sequence of substitutions may be infinite, we
represent it not as a \ct{LinkedList} object but as a \ct{Chain} object;
as usual, we find \ct{BlockClosure} objects good candidates to represent (a possibly
infinite) computation. Some examples are in order, \sd{ChainTest -> ChainExamples}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> collatz: o
    ^ o
        cons: [o even
                ifTrue: [self collatz: o / 2]
                ifFalse: [self collatz: 3 * o + 1]]

@@

ChainTest >> zeroOneTwo
    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
\end{minted}
where any \ct{Object} object respond to the message \ct{cons:} as follows: 
\sd{what is the argument?}
\sd{Separate Cons and Chain explanation}
\sd{why not just started with Cons. Ok we have an object}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> cons: anObj
    ^ anObj consedObject: self
\end{minted}
which dispatches over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> consedObject: car
    ^ Cons car: car asCons cdr: self asCons

BlockClosure >> consedObject: car
    ^ Chain item: car linker: self

Chain >> consedObject: car
    ^ car cons: [self]
\end{minted}
where (i)~\ct{Cons} objects mimic \textit{Lisp's cons cells}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons class >>  car: anObj cdr: anotherObj
    ^ self new
        car: anObj;
        cdr: anotherObj;
        yourself

Object >> asCons
    ^ Cons fromObject: self

Cons class >> fromObject: anObj
    ^ anObj
\end{minted}

\sd{I do not get fromObject: implementation. because you do not wrap the object.
So 1 asCons will return 1}
and (ii)~the abstract class \ct{Chain} provides the following class-side messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class >> bottom
    ^ Bottom new

Chain class >> item: anObj linker: aBlockClosure
    ^ Knot new
        item: anObj;
        linker: aBlockClosure;
        yourself

Chain class >> with: anObj
    ^ anObj cons: Chain bottom

Chain class >> repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
having two subclasses \ct{Knot} and \ct{Bottom}. A comprehensive
test case follows \sd{split it and explain.}
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \ct{Chain} object (i)~receives the message
\ct{next:} that carries a \ct{SmallInteger} object $n$ then it keeps at most
$n$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> next: n
    ^ self

@@

On the other hand, when a \ct{Chain} object (ii)~receives the message
\ct{\#contents} it builds a \ct{LinkedList} object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> contents
    ^ LinkedList new

Knot >> contents
    ^ self next contents
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot>>#next
    ^ linker value
\end{minted}
Observe that the \ct{Chain} hierarchy supports the protocol provided
by \ct{ReadStream} objects because of the messages \ct{\#next}, \ct{\#next:}
and \ct{\#contents}.

\subsection{\texttt{Runner}s, \texttt{Reifier}s and \texttt{Var}s}

Recall \sd{it was not said that way so I do not recall} that a \ct{Goal} object responds with a \ct{Chain} that contains either zero,
one, or more substitutions when it receives the message \ct{\#onState:} and it
delegates the task to process those substitutions to \ct{Runner} objects,

\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal >> solutions
    ^ Runner onGoal: self

Runner class >> onGoal: aGoal
    ^ self new
        goal: aGoal;
        yourself
\end{minted}
which are able to process both (i)~all of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> all
    ^ self run contents
\end{minted}
and (ii)~at most $n$ of them with the message \ct{atMost:} below: 
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> atMost: anInteger
    ^ (self run next: anInteger) contents
\end{minted}
relying on the unary message \ct{run}

\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner>>#run
    | states vars selector |
    states := goal onState: Dictionary new asState.
    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
    ^ states
        collect: [:s | selector
                        value: (s reifier reifyVars: vars)]
\end{minted}
having the responsibilities
\begin{itemize}

\item to send the message \Verb|#onState:| to \Verb|Goal| objects to start the
resolution process from an empty state, this can be considered the entry point of
the ``computation'' that produces a \Verb|Chain| of substitutions;

\item to take into account the variables with respect to results should be
provided; as a particular case, if the root goal doesn't introduce any logic
variable, by means of the auxiliary messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#vars
    ^ {}

Fresh>>#vars
    ^ vars
\end{minted}
then it is equivalent to prove a \textit{tautology}, represented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Var class>>#tautology
    ^ #tautology
\end{minted}
which require to decide that a relation is always true regardless values taken
by logic variables;

\item to simplify the results presentation when the root \Verb|Goal| object
introduces \textit{exactly one} logic variable: in this case a
\Verb|LinkedList| of values for that variable will be produced; otherwise, a
\Verb|LinkedList| of \Verb|LinkedList| of values will be produces each one for
each tuple of logic variables introduced by the root \Verb|Goal| object -- this
point will be clear when we'll explain \Verb|Fresh| goals that are the only
ones able to introduce logic variables;

\item to map each substitution (always translate to \Verb|State| object) that
satisfies the root \Verb|Goal| object do \textit{reify} the required variables,
this is accomplished by the messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
s reifier reifyVars: vars
\end{minted}
within the block argument of the returning \Verb|#collect:| message send.
\end{itemize}
The last point shows that \Verb|Chain| objects are \textit{functors} according
to the Haskell terminology because they answer to the message \Verb|#collect:|,
which is a synonism for the function \Verb|fmap :: (a -> b) -> f a -> f b|, as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain>>#collect: aBlockClosure
    ^ self

Knot>>#collect: aBlockClosure
    | v |
    v := aBlockClosure value: item.
    ^ v cons: [self next collect: aBlockClosure]
\end{minted}
respectively.  We are now ready to describe the reification process which
requires
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reifier
    ^ Reifier on: self

Reifier>>#reifyVars: vars
    | walker reifier |
    walker := state walker.
    reifier := Dictionary new asState reifier.
    ^ vars
        collect: [:v |
                    | w s |
                    w := walker walkStar: v.
                    s := reifier reify: w.
                    s walker walkStar: w]
\end{minted}
which uses the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier>>#reify: anObj 
    ^ (state walk: anObj) 
        reifyUsing: self
\end{minted}
to dispatch the message \Verb|#reifyUsing:| over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#reifyUsing: aReifier
    ^ aReifier reifyObject: self

Var>>#reifyUsing: aReifier 
    ^ aReifier reifyVar: self

ReifiedVar>>#reifyUsing: aReifier 
    ^ aReifier reifyReifiedVar: self

Array>>#reifyUsing: aReifier
    ^ aReifier reifyArray: self

LinkedList>>#reifyUsing: aReifier 
    ^ aReifier reifyLinkedList: self

Cons>>#reifyUsing: aReifier 
    ^ aReifier reifyCons: self 
\end{minted}
and to collect replies as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier>>#reifyObject: anObj 
    ^ state

Reifier>>#reifyVar: aVar 
    ^ state reifyVar: aVar

Reifier>>#reifyReifiedVar: aVar 
    ^ state

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyArray: aCollection 
    ^ self reifyCollection: aCollection

Reifier>>#reifyCollection: aCollection 
    ^ aCollection inject: state into: [:s :c | s reifier reify: c]
\end{minted}
requiring the auxiliary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#reifyVar: aVar 
    substitution
        at: aVar
        ifAbsentPut: [ReifiedVar new 
                        id: substitution size; 
                        yourself]
\end{minted}
and
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walk: anObj 
    | k |
    k := anObj.
    [k := substitution at: k ifAbsent: [^ k]] repeat
\end{minted}
Observe that \ct{Chain} objects are also \textit{foldable} according to the
Haskell \ct{Foldable} typeclass because they respond to the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons>>#inject: init into: aBlock 
    | accumulated |
    accumulated := aBlock value: init value: car.
    ^ aBlock value: accumulated value: cdr
\end{minted}

The reification process requires \ct{Walker} objects which are built by
\ct{State} objects using the template method
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State>>#walker
    ^ Walker on: self
\end{minted}
and it responds to the selector \Verb|#walkStar| with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStar: aVar
    ^ (state walk: aVar)
        walkStarUsingWalker: self
\end{minted}
and objects of different classes responds to \Verb|#walkStarUsingWalker:| as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarObject: self

Array>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarArray: self

LinkedList>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarLinkedList: self

Cons>>#walkStarUsingWalker: aWalker
    ^ aWalker walkStarCons: self
\end{minted}
dispatching over \Verb|Walker| objects
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker>>#walkStarObject: anObj
    ^ state walk: anObj

Walker>>#walkStarArray: aCollection
    ^ aCollection
        collect: [:each | self walkStar: each]

Walker>>#walkStarLinkedList: aCollection
    ^ aCollection
        collect: [:each | self walkStar: each]

Walker>>#walkStarCons: aCons
    | car cdr |
    car := self walkStar: aCons car.
    cdr := self walkStar: aCons cdr.
    ^ car cons: cdr
\end{minted}
respectively.

\section{Bootstrapping $\mu$Kanren}

@@

Let us consider the object \Verb|false| whose responsibility is to represent
\textit{falsehood}, hence it should be a good candidate to provide us something
related to this from the \textit{relational} point of view,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
false asGoal
\end{minted}
namely, \Verb|false| can build for us an object that represents a relation that
cannot be satisfied at all responding to the message selector \Verb|asGoal| --
don't be scared by multiple names, we freely interchange the terms
\textit{goal} and \textit{relation} postponing to later sections the
description of their technical difference. Moreover, let us
assume that the infrastructure that actually builds results is given and implemented
by the block \texttt{B := [:g | g solutions all]}, we write our first test case:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest>>#testFail
    self
        assert: false asGoal solutions all
        equals: {}
\end{minted}
Now we extend the \Verb|False| class with the following \Verb|asGoal| message
definition
\begin{minted}[fontsize=\footnotesize]{smalltalk}
False>>#asGoal
    ^ Goal fail
\end{minted}
and of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal>>#fail
    ^ Fail new
\end{minted}
such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
is the same to code the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail>>#onState: aState
    ^ Chain bottom
\end{minted}
which will be used within the message \Verb|solutions| sent in the block
\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
streams in particular, all in what follows.

>>>>

\sd{Why don't you start with Cons. Then once people understand Cons and infinite stream
you can introduce Goal and state and always example first.
\mn{I do prefer to explain both of them when the need arises to not postpone the main definition of Goal.}}

@@

In this section we describe the fundamentals of $\mu$Kanren and the protocols
that objects provide to compose the architecture on which Section
\ref{sec:bootstrapping} will define core logic operators.

We start describing objects that (i)~encodes relational statements,
(ii)~logical variables and (iii)~associations of values to variables such that
a given predicate can be satisfied.

\subsection{Goals, variables and states}

A \textit{goal} $g$ denotes a logical relation; precisely, $g$ is any object
that responds to the message \ct{\#onState:} so that it receives a substitution
and produces an ordered sequence of zero or more substitutions.

In turn, a substitution is an unordered collection of
values which are indexed by \textit{logic variables}, therefore a
\ct{Dictionary} object fits perfectly. For the sake of clarity, the predicate
\begin{displaymath}
inc(X, succ(succ(zero))) \quad\text{where}\quad Y = succ(X) \rightarrow inc(X, Y),
\end{displaymath}
is satisfied by the substitution
\begin{minted}[fontsize=\footnotesize]{smalltalk}
|X|
X := (Var id: 0).
Dictionary new
    at: X put: 'succ(zero)';
    yourself
\end{minted}

Moreover, logic variables are created by means of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Var class >> id: anInteger
    ^ self new
        id: ('var_' , anInteger asString) asSymbol;
        yourself
\end{minted}

We wrap a substitution in a \ct{State} object that keeps record of metadata
about the computation, for instance a \ct{State} has an integer \ct{birthdate}
to track the (discrete) time when a logic variable has been introduced -- its
meaning will be clarified later when we'll describe the \ct{Fresh}
goal\footnote{put here the reference to its subsection}.

An initial \ct{State} object is created by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary new asState
\end{minted}
where
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Dictionary >> asState
    ^ State new
        birthdate: 0;
        substitution: self;
        yourself
\end{minted}
and it can be augmented by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> at: aVar put: aValue
    | s |
    s := substitution copy.
    "check that if a key already exists at aVar, then the values must be the same"
    s
      at: aVar
      ifPresent: [:v | aValue = v ifFalse: [UnificationError signal]]
      ifAbsentPut: aValue.
    ^ self class new
        birthdate: birthdate;
        substitution: s;
        yourself
\end{minted}
maintaining a coherent protocol with respect to the one of \Verb|Dictionary|
objects for extension; finally, \ct{UnificationError} objects will be
explained in a later section when will talk about the unification process.

\subsection{Cons cells}

Because the sequence of substitutions may be infinite, we represent it not as a
\ct{LinkedList} object but as a \ct{Chain} object which has the capability to
represent potentially infinite sequences.  For this we will start building
\ct{Cons} cells; following the Lisp tradition, we reproduce the usual behavior
as the next test method shows:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
testCons
    self
        assert: '(0)'
        equals: (0 cons: nil) asString.
    self
        assert: '(0 1 2 3 4 5 6 7 8 9)'
        equals: (0 to: 9) asCons asString.
    self
        assert: '((1 2) 3 4 (5 (6 8)) 7)'
        equals: {{1. 2}. 3. 4. {5. {6. 8}}. 7} asCons asString.

    "dotted pairs notation"
    self
        assert: '(0 . 1)'
        equals: (0 cons: 1) asString.
    self
        assert: '(0 1 2 3 4 5 6 7 8 9 . 10)'
        equals: ((0 to: 9) asConsWithCdr: 10) asString.
    self
        assert: '(((3 . 4) . 5) 6 7)'
        equals: (((3 cons: 4) cons: 5) cons: (6 cons: (7 cons: nil))) asString
\end{minted}
We split its description of
\begin{itemize}
\item the creation message \ct{\#cons:} is defined as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> cons: anObj
    ^ anObj consedObject: self
\end{minted}
which dispatches the message \ct{\#consedObject:} on its argument to build
\ct{Cons} cells with
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> consedObject: car
    ^ Cons car: car cdr: self

Cons class >>  car: anObj cdr: anotherObj
    ^ self new
        car: anObj;
        cdr: anotherObj;
        yourself
\end{minted}

\item the translation message \ct{\#asCons} is defined and overridden as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> asCons
    ^ Cons fromObject: self

ByteSymbol >> asCons
    ^ Cons fromByteSymbol: self

SequenceableCollection >> asCons
    ^ Cons fromSequenceableCollection: self
\end{minted}
in parallel, the class object \ct{Cons} responds as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons >> fromObject: anObj
    ^ anObj

Cons >> fromByteSymbol: aSymbol
    ^ aSymbol

Cons >> fromSequenceableCollection: aCollection
    ^ self fromSequenceableCollection: aCollection lastCdr: nil

Cons >> fromSequenceableCollection: aCollection lastCdr: aCdr
    ^ aCollection
        foldr: [:each :acc | self car: each asCons cdr: acc asCons]
        init: aCdr
\end{minted}
Finally, to build the more general notion of \textit{dotted pairs},
the message \ct{\#asConsWithCdr:} is defined and overridden as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> asConsWithCdr: aCdr
    ^ Cons fromObject: self lastCdr: aCdr

SequenceableCollection >> asConsWithCdr: aCdr
    ^ Cons fromSequenceableCollection: self lastCdr: aCdr
\end{minted}
which require
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons >> fromObject: anObj lastCdr: aCdr
    ^ self car: anObj cdr: aCdr
\end{minted}

\end{itemize}

\subsection{Chains}

We find \ct{BlockClosure} objects good candidates to
represent a suspended (and possibly infinite) computation. The following
auxiliary methods for the test case \ct{ChainTest} show examples of how
\ct{BlockClosure}s can be used as arguments for the message \ct{\#cons:},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> collatz: o
    ^ o
        cons: [o even
                ifTrue: [self collatz: o / 2]
                ifFalse: [self collatz: 3 * o + 1]]

@@

ChainTest >> zeroOneTwo
    ^ 0 cons: (1 cons: (2 cons: Chain bottom))
\end{minted}
In turn, when a \ct{BlockClosure} receives the dispatched message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
BlockClosure >> consedObject: car
    ^ Chain item: car linker: self
\end{minted}
then it builds a \ct{Chain}; precisely, \ct{Chain} is an abstract class with
two subclasses \ct{Bottom} which denotes an empty sequence and \ct{Knot} which
denotes a sequence that has a value and a tail sequence that is not evaluated
yet, providing a \ct{BlockClosure} to do that, one item at a time.
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class >> bottom
    ^ Bottom new

Chain class >> item: anObj linker: aBlockClosure
    ^ Knot new
        item: anObj;
        linker: aBlockClosure;
        yourself
\end{minted}
Moreover, a \ct{Chain} can be used as argument of a \ct{\#cons:} message send
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain >> consedObject: car
    ^ car cons: [self]
\end{minted}
directly and other constructors are provided by \ct{Chain} class messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain class >> with: anObj
    ^ anObj cons: Chain bottom

Chain class >> repeat: anObj
    ^ anObj cons: [self repeat: anObj]
\end{minted}
that allow us to write the following comprehensive test case
\begin{minted}[fontsize=\footnotesize]{smalltalk}
ChainTest >> testNumbers
    self
        assert: Chain bottom contents
        equals: LinkedList new.
    self
        assert: (self nats next: 10) contents
        equals: (0 to: 9).
    self
        assert: (self zeroOneTwo next: 10) contents
        equals: (0 to: 2).
    self
        assert: (self zeroOneTwo next: 1) contents
        equals: (0 to: 0).
    self
        assert: (self fibs next: 10) contents
        equals: {0. 1. 1. 2. 3. 5. 8. 13. 21. 34}.
    self
        assert: ((self collatz: 10) next: 10) contents
        equals: {10. 5. 16. 8. 4. 2. 1. 4. 2. 1}.
    self
        assert: ((self gcd: 18 gcd: 32) next: 10) contents
        equals: {18. 32. 18. 14. 4. 2. 2}.
    self
        assert: ((Chain repeat: 4) next: 10) contents
        equals: {4. 4. 4. 4. 4. 4. 4. 4. 4. 4}
\end{minted}

On one hand, when a \ct{Chain} object receives the message
\ct{next:} that carries a \ct{SmallInteger} object $n$ then it keeps at most
$n$ objects and then truncate itself, precisely
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> next: n
    ^ self

@@

On the other hand, when a \ct{Chain} object receives the message
\ct{\#contents} it builds a \ct{LinkedList} object \textit{eagerly},
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Bottom >> contents
    ^ LinkedList new

Knot >> contents
    ^ self next contents
        addFirst: item;
        yourself
\end{minted}
where \Verb|Knot| objects provide the following simple but fundamental message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Knot >> next
    ^ linker value
\end{minted}
that allows us to do one more step in filling the not yet evaluated rest of the
sequence. Observe that the \ct{Chain} hierarchy supports the protocol provided
by \ct{ReadStream} objects because of the messages \ct{\#next}, \ct{\#next:}
and \ct{\#contents}.

\subsection{The Runner}

We can now be more precise and rephrase that a \ct{Goal} object responds with a
\ct{Chain} that contains either zero, one or more substitutions when it
receives the message \ct{\#onState:}.

In this section we describe three components that have the responsibility to
report and to present the substitutions that satisfy the given goal. The entry
point is the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal >> solutions
    ^ Runner onGoal: self

Runner class >> onGoal: aGoal
    ^ self new
        goal: aGoal;
        yourself
\end{minted}
that allows a goal to build a \ct{Runner} which has the responsibilities to
process the \ct{Chain} of substitutions; in particular, (i)~all of them with
the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> all
    ^ self run contents
\end{minted}
and (ii)~at most $n$ of them with the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> atMost: anInteger
    ^ (self run next: anInteger) contents
\end{minted}

Those messages rely on the unary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Runner >> run
    | states vars selector |
    states := goal onState: Dictionary new asState.
    vars := goal vars ifEmpty: [LinkedList with: Var tautology].
    selector := vars size = 1 ifTrue: #first ifFalse: #yourself.
    ^ states
        collect: [:s | selector
                        value: (s reifier reifyVars: vars)]
\end{minted}
that accomplishes the responsibilities
\begin{itemize}

\item to send the message \Verb|#onState:| to a \Verb|Goal| $g$ in order to
start the resolution process from an empty state -- this can be considered the
entry point of the ``computation'' that produces the \Verb|Chain| of
substitutions that satisfy $g$;

\item to take into account the variables with respect to results should be
provided; as a particular case, if the root \ct{Goal} $g$ doesn't introduce any logic
variable, recovered by the messages
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal >> vars
    ^ {}

Fresh >> vars
    ^ vars
\end{minted}
then it is equivalent to prove that $g$ is a \textit{tautology}, represented
by
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Var class >> tautology
    ^ #tautology
\end{minted}
which requires to decide that a relation is always \textit{true} regardless
the values associated to logic variables;

\item to simplify the presentation of results when the root \Verb|Goal| object
introduces \textit{exactly one} logic variable $v$: in this case a
\Verb|LinkedList| of values associated to $v$ will be produced; otherwise, a
\Verb|LinkedList| of \Verb|LinkedList| of values will be produced, a list for
each tuple of logic variables introduced -- this point will become clear when
we'll explain \Verb|Fresh| goals that are the only ones able to introduce logic
variables;

\item to map each substitution (always translate to \Verb|State| object) that
satisfies the root \Verb|Goal| object to a \textit{reified} substitution.
\end{itemize}

The last point shows that \Verb|Chain| objects are \textit{functors} according
to the Haskell terminology because they answer to the message \Verb|#collect:|,
which is a synonism for the function \Verb|fmap :: (a -> b) -> f a -> f b|, as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Chain >> collect: aBlockClosure
    ^ self

Knot >> collect: aBlockClosure
    | v |
    v := aBlockClosure value: item.
    ^ v cons: [self next collect: aBlockClosure]
\end{minted}
respectively.

\subsection{The Reifier}

We are now ready to describe the reification process which
requires
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> reifier
    ^ Reifier on: self

Reifier >> reifyVars: vars
    | walker reifier |
    walker := state walker.
    reifier := Dictionary new asState reifier.
    ^ vars
        collect: [:v |
                    | w s |
                    w := walker value: v.
                    s := reifier value: w.
                    s walker value: w]
\end{minted}
which uses the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier >> value: anObj
    ^ (state walk: anObj)
        reifyUsingReifier: self
\end{minted}
to dispatch the message \Verb|#reifyUsingReifier:| over
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> reifyUsingReifier: aReifier
    ^ aReifier reifyObject: self

Var >> reifyUsingReifier: aReifier
    ^ aReifier forVar: self

ReifiedVar >> reifyUsingReifier: aReifier
    ^ aReifier forReifiedVar: self

Array >> reifyUsingReifier: aReifier
    ^ aReifier forArray: self

LinkedList >> reifyUsingReifier: aReifier
    ^ aReifier forLinkedList: self

Cons >> reifyUsingReifier: aReifier
    ^ aReifier forCons: self
\end{minted}
and to collect replies as
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Reifier >> forObject: anObj
    ^ state

Reifier >> forVar: aVar
    ^ state reifyVar: aVar

Reifier >> forReifiedVar: aVar
    ^ state

Reifier >> forArray: aCollection
    ^ self _forCollection: aCollection

Reifier >> forLinkedList: aCollection
    ^ self _forCollection: aCollection

Reifier >> forCons: aCollection
    ^ self _forCollection: aCollection

Reifier >> _forCollection: aCollection
    ^ aCollection inject: state into: [:s :c | s reifier value: c]
\end{minted}
requiring the auxiliary message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> reifyVar: aVar
    substitution
        at: aVar
        ifAbsentPut: [ReifiedVar new
                        id: substitution size;
                        yourself]
\end{minted}
and
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> walk: anObj
    | k |
    k := anObj.
    [k := substitution at: k ifAbsent: [^ k]] repeat
\end{minted}
Observe that \ct{Chain} objects are also \textit{foldable} according to the
Haskell \ct{Foldable} typeclass because they respond to the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Cons >> inject: init into: aBlock
    | accumulated |
    accumulated := aBlock value: init value: car.
    ^ aBlock value: accumulated value: cdr
\end{minted}

\subsection{The Walker}
The reification process requires \ct{Walker} objects which are built by
\ct{State} objects using the template method
\begin{minted}[fontsize=\footnotesize]{smalltalk}
State >> walker
    ^ Walker on: self
\end{minted}
and it responds to the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker >> value: aVar
    ^ (state walk: aVar)
        walkStarUsingWalker: self
\end{minted}
and objects of different classes responds to \Verb|#walkStarUsingWalker:| as follows
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Object >> walkStarUsingWalker: aWalker
    ^ aWalker forObject: self

Array >> walkStarUsingWalker: aWalker
    ^ aWalker forArray: self

LinkedList >> walkStarUsingWalker: aWalker
    ^ aWalker forLinkedList: self

Cons >> walkStarUsingWalker: aWalker
    ^ aWalker forCons: self
\end{minted}
dispatching over \Verb|Walker| objects
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Walker >> forObject: anObj
    ^ state walk: anObj

Walker >> forArray: aCollection
    ^ self _forCollection: aCollection

Walker >> forLinkedList: aCollection
    ^ self _forCollection: aCollection

Walker >> forCons: aCons
    ^ self _forCollection: aCons

Walker >> _forCollection: aCollection
    ^ aCollection
        collect: [:each | self value: each]
\end{minted}
respectively.

\section{Bootstrapping $\mu$Kanren}
\label{sec:bootstrapping}

@@

Let us consider the object \Verb|false| whose responsibility is to represent
\textit{falsehood}, hence it should be a good candidate to provide us something
related to this from the \textit{relational} point of view,
\begin{minted}[fontsize=\footnotesize]{smalltalk}
false asGoal
\end{minted}
namely, \Verb|false| can build for us an object that represents a relation that
cannot be satisfied at all responding to the message selector \Verb|asGoal| --
don't be scared by multiple names, we freely interchange the terms
\textit{goal} and \textit{relation} postponing to later sections the
description of their technical difference. Moreover, let us
assume that the infrastructure that actually builds results is given and implemented
by the block \texttt{B := [:g | g solutions all]}, we write our first test case:
\begin{minted}[fontsize=\footnotesize]{smalltalk}
GoalTest >> testFail
    self
        assert: false asGoal solutions all
        equals: {}
\end{minted}
Now we extend the \Verb|False| class with the following \Verb|asGoal| message
definition
\begin{minted}[fontsize=\footnotesize]{smalltalk}
False >> asGoal
    ^ Goal fail
\end{minted}
and of the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Goal >> fail
    ^ Fail new
\end{minted}
such that to say that \verb|Fail| denotes the impossibility to satisfy a relation
is the same to code the message
\begin{minted}[fontsize=\footnotesize]{smalltalk}
Fail >> onState: aState
    ^ Chain bottom
\end{minted}
which will be used within the message \Verb|solutions| sent in the block
\Verb|B| -- a lot remains to be explained, such as the \Verb|Chain| class and
streams in particular, all in what follows.
